# Week 6: Hypothesis Testing
You've made it to the final lab of this course! This week is all about hypothesis testing, one of the most ubiquitous elements of statistical theory, and the conclusion to your journey into inferential statistics (for now...). Almost any quantitative research paper you will read will feature a hypothesis test in one form or another. Furthermore, the quantitative research you might conduct in other courses and/or your capstone will rely heavily on the framework hypothesis testing. During this lab we will first discuss the theories at the foundation of hypothesis testing. In the second part of the lab we are going to practice with conducting null hypothesis significance tests (NHSTs) in R. 

To get started, download the [lab template here](https://github.com/thomashulst/quantrma_lab/raw/master/Labs/Lab06_HypothesisTesting.Rmd) (right click: save as) or from Canvas. Copy the lab template to your lab folder and double click the **Lab project file** (not the .Rmd) to open RStudio. 

## Learning goals
During this lab you will do the following:

1. Discuss various concepts related to hypothesis testing
2. Conduct one-sample, paired-samples and independent samples $t$-tests in R

## Research methods: hypothesis testing

### Question 1
What is a **research** hypothesis? What is a **statistical** hypothesis? What is the difference between the two? Explain in your own words.

### Question 2
What is the null and alternative hypothesis of the following cases:

#### Case 1
You flip a coin 50 times and count the number of heads flipped ($X = 12$). Suppose you want to conduct a hypothesis test to determine if the underlying probability of flipping heads for this particular coin is equal to $0.30$. 

#### Case 2
A random survey of 75 death row inmates revealed that the mean length of time on death row is 17.4 years with a standard deviation of 6.3 years. Suppose you want to conduct a hypothesis test to determine if the mean time on death row is 15 years.^[Case from [*Introductory Statistics*](https://opentextbc.ca/introstatopenstax/front-matter/preface/) by OpenStax]

#### Case 3
You are a researcher investigating the effect of age on working memory. You have collected data on how many random digits can be remembered for a brief period of time in a group of 20 young adults ($\bar{X} = 7.2$ digits) and 20 elderly participants ($\bar{X} = 6.9$ digits). The standard deviation is 1 digit in both groups. Suppose you want to conduct a hypothesis test to determine if young adults have a *better* working memory than elderly.

#### Case 4
A course coordinator is interested if their course about statistics actually has an effect on the statistical ability of students. The course coordinator has 200 students take a quiz about statistics before the course starts (mean pre-course quiz grade = 50 points) and has the same students take a quiz after completion of the course (mean post-course quiz grade = 75 points). The course coordinator made sure the questions were different between the quizzes, but equivalent in difficulty. The standard deviation of the difference is 10 points. Suppose the course coordinator wants to conduct a hypothesis test to determine if the course had *any* effect on the statistical ability of students (i.e. quiz grades getting better or worse).

### Question 3
What is the appropriate statistical test to conduct for each of the statistical hypotheses stated in Question 2? Explain why a particular statistical test is appropriate. Assume the population standard deviation is unknown for case 2, 3 and 4. 

### Question 4
What is meant by the critical region for a test statistic? What does it mean if the test statistic of your sample falls within the critical region? Explain in your own words. 

### Question 5
What is the meaning of the p-value of a statistical test? Explain in your own words.

### Question 6
The following is the output of the statistical tests which tested the hypotheses of the cases stated in Question 2. Comment on the results of each statistical test. Do you reject or retain the null hypothesis in each case?

#### Case 1

    > binom.test(X,N,theta)
    
    Exact binomial test
    
    data:  X and N
    number of successes = 12, number of trials = 50, p-value = 0.4407
    alternative hypothesis: true probability of success is not equal to 0.3
    95 percent confidence interval:
      0.1306099 0.3816907
    sample estimates:
    probability of success 
      0.24 


#### Case 2

    > t.test(time_deathrow, mu = 15)
    
    One Sample t-test
    
    data:  time_deathrow
    t = 3.2991, df = 74, p-value = 0.001493
    alternative hypothesis: true mean is not equal to 15
    95 percent confidence interval:
      15.9505 18.8495
    sample estimates:
    mean of x 
     17.4 

#### Case 3

    > t.test(young_adults,elderly,alternative = "greater")

    Welch Two Sample t-test
    
    data:  young_adults and elderly
    t = 0.94868, df = 38, p-value = 0.1744
    alternative hypothesis: true difference in means is greater than 0
    95 percent confidence interval:
      -0.2331456        Inf
    sample estimates:
    mean of x mean of y 
      7.2       6.9 

#### Case 4

    > t.test(pre,post,paired =T)
    
    Paired t-test

    data:  pre and post
    t = -80.58, df = 199, p-value < 2.2e-16
    alternative hypothesis: true difference in means is not equal to 0
    95 percent confidence interval:
      -25.6118 -24.3882
    sample estimates:
    mean of the differences 
      -25


### Question 7
What does it mean if the result of a statistical test is significant? Does statistical significance say anything about whether the result is important? 

### Question 8
What type of error is committed in the following cases? Which type of error has a greater consequence in this particular case?

#### Case 1
You are a beginning rock climber and have bought some second hand climbing equipment on Marktplaats. Assume your null hypothesis about the safety of the rock climbing equipment is: the climbing equipment is safe. What type of error do you commit if you think the rock climbing equipment is not safe, when in fact it is safe? 

#### Case 2
You are a beginning rock climber and have bought some second hand climbing equipment on Marktplaats. Assume your null hypothesis about the safety of the rock climbing equipment is: the climbing equipment is safe. What type of error do you commit if you think the rock climbing equipment is safe, when in fact it is not safe? 

### Question 9
What does the power of a statistical test refer to? Explain in your own words.

## $t$-tests in R
I'll be honest. Conducting a $t$-test in R can be slightly painful. Part of that has to do with the fact that the `t.test()` function was not really written with "tidy data" in mind (which we have become used to over the past couple of weeks), but also because it can do more than just one thing. It can actually do three different types of $t$-test (more even, but we focus on the three here) so the function is quite specific in how it expects its arguments depending on which test you want it to perform. How to use `t.test()` is actually explained quite extensively [in the textbook](https://thomashulst.github.io/quantrma/week-6-hypothesis-testingweek6-hypothesistesting-1.html#doing-the-test-in-r), so I will not go over all the details here, but I will highlight some useful tips while conducting $t$-tests in R. 

Remember the three types of $t$-test we can conduct are: the one-sample $t$-test, the paired samples $t$-test and the independent samples $t$-test. The one-sample $t$-test is simplest to use in R. Suppose we want to compare the mean grade of one group of students to a hypothesized mean. In that case, `t.test()` expects two arguments: `x` (a vector with your data) and `mu` (the value of the null hypothesis).  

```{r,echo=T}
set.seed(123)
grades <- rnorm(n = 10, 75, 5) # generate 10 grades
null <- 60 # value of null hypothesis

results <- t.test(x = grades, mu = null)

print(results)
```

We have saved the results of the $t$-test in a variable called `results`. If we want to work with a specific result (for example the confidence interval), we can do that as follows:

```{r,echo=T}

results$conf.int

```

More will follow shortly.

## Stroop data
We are going to have another look at the Stroop experiment data from last week. The Stroop data should still be saved in your lab folder, so load the data as we did during the last lab:

```{r}
library(readxl)
# read_xlsx depends on where you have placed and how you have named your file
df_stroop_wide <- read_xlsx("data/stroop.xlsx") 
df_stroop_long <- pivot_longer(df_stroop_wide,cols = c(Congruent,Incongruent),names_to = "Condition",values_to = "RT")
```

If you don't have a copy of last week's data saved, you can download an example file from Canvas under the lecture for Week 6.

### $t$-test exercises
A colleague hypothesizes that the average reaction time for the Congruent condition is 1000ms in the population ($\mu = 1000$). Suppose you want to test if there is a statistically significant difference between the mean reaction time in the Congruent condition of your sample and the mean of the population.

1. State the null and alternative hypothesis for this test.
2. Conduct the appropriate statistical test to test these hypotheses.
3. What do you conclude about the null hypothesis? 

We continue working with the Stroop data from last week, but this time we are interested in testing if there is a statistically significant difference in reaction time between the Congruent and Incongruent condition. We strongly expect a positive difference and only want to test positive Stroop effects (i.e. Incogruent - Congruent > 0). Complete the following exercises: 

1. State the null and alternative hypothesis for this test. 
2. Conduct the appropriate statistical test to test these hypotheses.
3. What do you conclude about the null hypothesis? 

Finally, we are going to finish this lab by do something "illegal". Don't worry, you aren't going to get arrested, but I do want you to understand the gravity of the statistical crime we are going to commit. It's for a good reason, so we are allowing it for once...

We know the data in the Stroop experiment was collected in pairs: you took part in both the Congruent and Incongruent condition. However, how would we analyze the Stroop data if we disregard this fact and pretend the data was collected in independent groups instead? So, let's pretend, for didactic purposes, the reaction times for the Congruent condition were collected in one group of students and the Incongruent condition in a different group of students. 

1. State the null and alternative hypothesis for this test. 
2. Conduct the appropriate statistical test to test these hypotheses.
3. What do you conclude about the null hypothesis? 

When you have completed all exercises and are happy with your progress today, please knit your document and submit it to Canvas. If you finish before the time is up, think about what topics from this course you would like to revise, work on your assignment, or help out your fellow students.
