---
title: 'Tutorial 1 - Introduction'
author: "Student Name"
date: "xx/xx/xx"
output: 
  html_document:
    toc: true
    toc_float: true

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

# Research methods exercises (first part)
**Answer question 1**

Answer here.

**Answer question 2**

Answer here.

**Answer question 3**

Answer here.

**Answer question 4**

Answer here.

**Answer question 5**

Answer here.

**Answer question 6**

Answer here.

**Answer question 7**

Answer here.


# Tutorial manual examples R (second part)

You can keep notes about the R part of the tutorial here and copy and paste R code in the code block below.

```{r, echo=TRUE,eval=TRUE}

# Copy and paste the R code of the examples from the tutorial manual here.

```

# Generalization exercises R (second part)


```{r, echo=TRUE,eval=TRUE}

# Use the code from the tutorial manual examples to attempt to solve this assignment

#1. Make a graph plotting Life Expectancy by year for the five continents, using the `continent` factor. Make sure you change the title as well.

#2. Make a graph plotting GDP per capita by year for the USA, Canada, and Mexico. Use the `gdpPercap` column for the GDP per capita data.

#3. Make a new graph plotting anything you are interested in using the gapminder dataset. It just needs to be a plot that we have not given an example for.

```

# Writing assignment R (second part)

Describe what histograms are, how to interpret them, and what they are useful for. You should answer each of these questions:

a. What do the bars on a histogram represent?
b. How many bars can a histogram have?
c. What do the heights of the bars tell you
d. What is on the x-axis and y-axis of a histogram
e. What does the tallest bar on a histogram tell you?
f. What does the shortest bar on a histogram tell you?
g. What are some uses for histograms, why would you want to look at a histogram of some numbers that you collected?
h. Imagine you had two histograms, one was very wide and spread out, the other was very narrow with a very tall peak. Which histogram would you expect to contain more consistent numbers (numbers that are close to each other) and explain why.


