# Week 2: Describing Data

>Describing comic sensibility is near impossible. It's sort of an abstract silliness, that sometimes the joke isn't the star.
>---Dana Carvey

The purpose of this lab is to show you how to generate graphs and compute basic descriptive statistics, including measures of central tendency (mean, mode, median) and variation (range, variance, standard deviation). 

## General Goals

1. Learn how to load data into R
2. Learn how data is structured
3. Compute measures of central tendency using R
4. Compute measures of variation using R
5. Ask some questions of a data set using descriptive statistics

## Descriptives basics in R

We learned in lecture and from the textbook that data we want to use ask and answer questions often comes with loads of numbers. Too many numbers to look at all at once. That's one reason we use descriptive statistics. To reduce the big set of numbers to one or two summary numbers that tell use something about all of the numbers. R can produce descriptive statistics for you in many ways. There are base functions for most of the ones that you want. We'll go over some R basics for descriptive statistics, and then use our new found skills to ask some questions about real data.

### Making numbers in R

In order to do descriptive statistics we need to put some numbers in a variable. You can also do this using the `c()` command, which stands for combine

```{r}
my_numbers <- c(1,2,3,4)
```

There a few other handy ways to make numbers. We can use `seq()` to make a sequence. Here's making the numbers from 1 to 100

```{r}
one_to_one_hundred <- seq(1,100,1)
```

We can repeat things, using rep. Here's making 10 5s, and 25 1s:

```{r}
rep(10,5)
rep(1,25)
all_together_now <- c(rep(10,5),rep(1,25)) 
```

#### Sum

Let's play with the number 1 to 100. First, let's use the `sum()` function to add them up

```{r}
one_to_one_hundred <- seq(1,100,1)
sum(one_to_one_hundred)
```

#### Length

We put 100 numbers into the variable `one_to_one_hundred`. We know how many numbers there are in there. How can we get R to tell us? We use `length()` for that. 

```{r}
length(one_to_one_hundred)
```

### Central Tendency

#### Mean

Remember the mean of some numbers is their sum, divided by the number of numbers. We can compute the mean like this:

```{r}
sum(one_to_one_hundred)/length(one_to_one_hundred)
```

Or, we could just use the `mean()` function like this:

```{r}
mean(one_to_one_hundred)
```

#### Median

The median is the number in the exact middle of the numbers ordered from smallest to largest. If there are an even number of numbers (no number in the middle), then we take the number in between the two (decimal .5). Use the `median` function. There's only 3 numbers here. The middle one is 2, that should be the median

```{r}
median(c(1,2,3))
```

#### Mode

R does not a base function for the Mode. You would have to write one for yourself. Here is an example of writing your own mode function, and then using it. Note I searched how to do this on Google, and am using the mode defined by [this answer on stack overflow](https://stackoverflow.com/questions/2547402/is-there-a-built-in-function-for-finding-the-mode)

Remember, the mode is the most frequently occurring number in the set. Below 1 occurs the most, so the mode will be one.

```{r}

my_mode <- function(x) {
  ux <- unique(x)
  ux[which.max(tabulate(match(x, ux)))]
}

my_mode(c(1,1,1,1,1,1,1,2,3,4))


```

### Variation

We often want to know how variable the numbers are. We are going to look at descriptive statistics to describe this such as the **range**, **variance**, the **standard deviation**, and a few others.

First, let's remind ourselves what variation looks like (it's when the numbers are different). We will sample 100 numbers from a normal distribution (don't worry about this yet), with a mean of 10, and a standard deviation of 5, and then make a histogram so we can see the variation around 10..

```{r}
sample_numbers <- rnorm(100,10,5)
hist(sample_numbers)
```

#### range

The range is the minimum and maximum values in the set, we use the `range` function.

```{r}
range(sample_numbers)
```

#### var = variance

We can find the sample variance using `var`. Note, divides by (n-1)

```{r}
var(sample_numbers)

```

#### sd = standard deviation

We find the sample standard deviation us SD. Note, divides by (n-1)

```{r}
sd(sample_numbers)
```

Remember that the standard deviation is just the square root of the variance, see:

```{r}
sqrt(var(sample_numbers))
```

#### All Descriptives

Let's put all of the descriptives and other functions so far in one place:

```{r}
sample_numbers <- rnorm(100,10,5)

sum(sample_numbers)
length(sample_numbers)
mean(sample_numbers)
median(sample_numbers)
my_mode(sample_numbers)
range(sample_numbers)
var(sample_numbers)
sd(sample_numbers)

```

### Descriptives by conditions

Sometimes you will have a single variable with some numbers, and you can use the above functions to find the descriptives for that variable. Other times (most often in this course), you will have a big data frame of numbers, with different numbers in different conditions. You will want to find descriptive statistics for each the sets of numbers inside each of the conditions. Fortunately, this is where R really shines, it does it all for you in one go. 

Let's illustrate the problem. Here I make a date frame with 10 numbers in each condition. There are 10 conditions, each labelled, A, B, C, D, E, F, G, H, I, J.

```{r}
scores <- rnorm(100,10,5)
conditions <- rep(c("A","B","C","D","E","F","G","H","I","J"), each =10)
my_df <- data.frame(conditions,scores)
```

If you look at the `my_df` data frame, you will see it has 100 rows, there are 10 rows for each condition with a label in the `conditions` column, and 10 scores for each condition in the `scores` column. What if you wanted to know the mean of the scores in each condition? You would want to find 10 means.

The slow way to do it would be like this:

```{r}
mean(my_df[my_df$conditions=="A",]$scores)
mean(my_df[my_df$conditions=="B",]$scores)
mean(my_df[my_df$conditions=="C",]$scores)
# and then keep going
```

Nobody wants to do that! Not, me I stopped doing it that way, you should to.

#### group_by and summarise

We can easily do everything all at once using the `group_by` and `summarise` function from the `dplyr` package. Just watch

```{r}
library(dplyr)

my_df %>%
  group_by(conditions) %>%
  summarise(means = mean(scores))

```

A couple things now. First, the print out of this was ugly. Let's fix that. we put the results of our code into a new variable, then we use `knitr::kable` to print it out nicely when we `knit` the document


```{r}

summary_df <- my_df %>%
               group_by(conditions) %>%
               summarise(means = mean(scores))

knitr::kable(summary_df)

```

#### multiple descriptives

The best thing about the `dplyr` method, is that we can add more than one function, and we'll get more than one summary returned, all in a nice format, let's add the standard deviation:

```{r}

summary_df <- my_df %>%
               group_by(conditions) %>%
               summarise(means = mean(scores),
                         sds = sd(scores))

knitr::kable(summary_df)

```

We'll add the min and the max too:

```{r}

summary_df <- my_df %>%
               group_by(conditions) %>%
               summarise(means = mean(scores),
                         sds = sd(scores),
                         min = min(scores),
                         max = max(scores))

knitr::kable(summary_df)

```

### Describing gapminder

Now that we know how to get descriptive statistics from R, we cam do this will some real data. Let's quickly ask a few question about the gapminder data. 

```{r}
library(gapminder)
gapminder_df <- gapminder
```

Note: The above code will only work if you have installed the gapminder package. Make sure you are connected to the internet, then choose the Packages tab from the bottom right panel, and choose install. Thens search for gapminder, choose it, and install it. 

#### What are some descriptive for Life expectancy by continent?

Copy the code from the last part of descriptives using `dplyr`, then change the names like this:

```{r}

summary_df <- gapminder_df %>%
               group_by(continent) %>%
               summarise(means = mean(lifeExp),
                         sds = sd(lifeExp),
                         min = min(lifeExp),
                         max = max(lifeExp))

knitr::kable(summary_df)

```

### Generalization Exercise

(1 point - Pass/Fail)

Complete the generalization exercise described in your R Markdown document for this lab.

1. What is the mean, standard deviation, minimum and maximum life expectancy for all the gapminder data (across all the years and countries). Hint: do not use `group_by`

2. What is the mean, standard deviation, minimum and maximum life expectancy for all of the continents in 2007, the most recent year in the dataset. Hint: add another pipe using `filter(year==2007) %>%`

### Writing assignment

(2 points - Graded)

Complete the writing assignment described in your R Markdown document for this lab. When you have finished everything. Knit the document and hand in your stuff (you can submit your .RMD file to blackboard if it does not knit.)

Your writing assignment is to answer these questions in full sentences using simple plain langauge:

1. Define the mode.
2. Explain what would need to happen in order for a set of numbers to have two modes
3. Define the median
4. Define the mean
5. Define the range
6. When calculating the standard deviation, explain what the difference scores represent
7. Explain why the difference scores are squared when calculating the standard deviation
8. If one set of numbers had a standard deviation of 5, and another had a standard deviation of 10, which set of numbers would have greater variance, explain why.


## Part two: Graphing data

>The commonality between science and art is in trying to see profoundly - to develop strategies of seeing and showing.
>---Edward Tufte

As we have found out from the textbook and lecture, when we measure things, we get lots of numbers. Too many numbers. Sometimes so many your head explodes just thinking about them. One of **the most helpful things** you can do to begin to make sense of these numbers, is to look at them in graphical form. Unfortunately, for sight-impaired individuals, graphical summary of data is much more well-developed than other forms of summarizing data for our human senses. Some researchers are developing auditory versions of visual graphs, a process called **sonification**, but we aren't prepared to demonstrate that here. Instead, we will make charts, and plots, and things to look at, rather than the numbers themselves, mainly because these are tools that are easiest to get our hands on, they are the most developed, and they work really well for visual summary. If time permits, at some point I would like to come back here and do the same things with sonification. I think that would be really, really cool!

During this part of the lab we'll do these things in R:

1. Load some data in R
2. Talk a little bit about how the data is structured
3. Make graphs of the data so we can look at it and make sense of it

### Important info

1. Data for NYC film permits was obtained from the NYC open data website. The raw .csv file can be found here: <a href="https://raw.githubusercontent.com/thomashulst/quantrma_lab/master/data/Film_Permits.csv" download>Film_Permits.csv</a>


2. Gapminder data from the gapminder project (copied from the R gapminder library) can be downloaded in raw .csv format here: <a href="https://raw.githubusercontent.com/thomashulst/quantrma_lab/master/data//gapminder.csv" download>gapminder.csv</a>

### Generate some data

In order to graph data, we need to have some data first...Actually, with R, that's not quite true. Run the following bit of code and see what happens. Note that from now on, any text in the grey blocks is R code which can be run by copy/pasting it in the R Studio console and pressing enter. Additionally, you can copy and paste the code below to your lab file so as to keep notes of your work. 

```{r}
hist(rnorm(100, mean=50, sd=25))
```

You just made R sample 100 numbers, and then plot the results in a histogram. Pretty neat. We'll be doing some of this later in the course, where get R to make fake data for us, and then we learn to think about how data behaves under different kinds of assumptions.

For now, let's do something that might be a little bit more interesting... let's figure out what movies are going to be filmed in NYC. It turns out that NYC makes a lot of data about a lot things open and free for anyone to download and look at. This is the NYC Open Data website: [https://opendata.cityofnewyork.us](https://opendata.cityofnewyork.us). I searched through the data, and found a data file that lists the locations of film permits for shooting movies all throughout the Burroughs. There are multiple ways to load this data into R.

1. If you have downloaded the [RMarkdownLabs.zip](https://github.com/thomashulst/quantrma_lab/blob/master/RMarkdownsLabs.zip?raw=true) file, then you already have the data file in the data folder. Assuming you are working in your main directory (your .rmd file is saved in the main folder that contains both the data and template folders), then use the following commands to load the data.

```{r, eval=TRUE}
library(data.table)
nyc_films <-fread("data/Film_Permits.csv")
```

2. If the above method doesn't work, you can try loading the data from the course website using:

```{r, eval=FALSE}
library(data.table)
nyc_films <- fread("https://raw.githubusercontent.com/thomashulst/quantrma_lab/master/data/Film_Permits.csv")
```

If you are having issues getting the data loaded, talk to your tutor.

### Look at the data

You will be downloading and analyzing all kinds of data files this quad. We will follow the very same steps every time. The steps are to load the data, then look at it. You want to see what you've got.

In R Studio, you will now see a variable called `nyc_films` in the top right-hand corner of the screen, in the environment tab. If you click this thing, it will show you the contents of the data in a new window. The data is stored in something we call a `data frame`. It's R lingo, for the thing that contains the data. Notice it is shaped like a rectangle, with rows going across, and columns going up and down. It looks kind of like an excel spreadsheet if you are familiar with Excel.

It's useful to know you can look at the data frame this way if you need to. But, this data frame is really big, it has 50,728 rows of data. That's a lot too much to look at.

#### summarytools

The summarytools package gives a quick way to summarize all of the data in a data frame. Here's how. When you run this code you will see the summary in the viewer on the bottom right hand side. There's a little browser button (arrow on top of little window) that you can click to expand and see the whole thing in a browser.

```{r}
library(summarytools)
view(dfSummary(nyc_films))
```

That is super helpful, but it's still a lot to look at. Because there is so much data here, it's pretty much mind-boggling to start thinking about what to do with it.

### Make plots to answer questions

Let's walk through a couple questions we might have about this data. We can see that there were 50,728 film permits made. We can also see that there are different columns telling us information about each of the film permits. For example, the `Borough` column lists the Borough for each request, whether it was made for: Manhattan, Brooklyn, Bronx, Queen's, or Staten Island. Now we can ask our first question, and learn how to do some plotting in R. 

#### Where are the most film permits being requested?

Do you have any guesses? Is it Manhattan, or Brooklyn, of the Bronx? Or Queen's or Staten Island? We can find out by plotting the data using a bar plot. We just need to count how many film permits are made in each borough, and then make different bars represent the the counts.

First, we do the counting in R. Run the following code:

```{r, eval=TRUE}
library(dplyr)

counts <- nyc_films %>%
          group_by(Borough) %>%
          summarize(count_of_permits = length(Borough))

```

The above grouped the data by each of the five Borough's, and then counted the number of times each Borough occurred (using the `length` function). The result is a new variable called `count`. I chose to name this variable `count`. You can see that it is now displayed in the top-right hand corned in the environment tab. If you gave `count` a different name, like `muppets`, then it would be named what you called it.

If you click on the `counts` variable, you will see the five boroughs listed, along with the counts for how many film permits were requested in each Borough. These are the numbers that we want to plot in a graph.

We do the plot using a fantastic package called `ggplot2`. It is very powerful once you get the hand of it, and when you do, you will be able to make all sorts of interesting graphs. Here's the code to make the plot:

```{r 1borough,echo=TRUE,eval=TRUE}
library(ggplot2)

ggplot(counts, aes(x = Borough, y = count_of_permits )) +
  geom_bar(stat="identity")

```


There it is, we're done here! We can easily look at this graph, and answer our question. Most of the film permits were requested in Manhattan, followed by Brooklyn, then Queen's, the Bronx, and finally Staten Island. 

#### What kind of "films" are being made, what is the category?

We think you might be skeptical of what you are doing here, copying and pasting things. Soon you'll see just how fast you can do things by copying and pasting, and make a few little changes. Let's quickly ask another question about what kinds of films are being made. The column `Category`, gives us some information about that. Let's just copy paste the code we already made, and see what kinds of categories the films fall into. See if you can tell what I changed in the code to make this work, I'll do it all at once:

```{r 1category,echo=TRUE,eval=TRUE}

counts <- nyc_films %>%
          group_by(Category) %>%
          summarize(count_of_permits = length(Category))

ggplot(counts, aes(x = Category, y = count_of_permits )) +
  geom_bar(stat="identity")

```

OK, so this figure might look a bit weird because the labels on the bottom are running into each other. We'll fix that in a bit. First, let's notice the changes. 

1. I changed `Borough` to `Category`. That was the main thing

2. I left out a bunch of things from before. None of the `library()` commands are used again, and I didn't re-run the very early code to get the data. R already has those things in it's memory, so we don't need to do that first. If you ever clear the memory of R, then you will need to reload those things.

Fine, so how do we fix the graph? Good question. You probably have no idea how to do this, but googling your questions is a great way to learn R. It's what everybody does.

When you Google this question, you will likely find a lot of ways to fix your graph. The trick I used is to add the last line in the R code below. I just copy-pasted it from the solution I found on [stack overflow](https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2) (you will become friend's with stack overflow, there are many solutions there to all of your questions).


```{r 1categoryB,echo=TRUE,eval=TRUE}

counts <- nyc_films %>%
          group_by(Category) %>%
          summarize(count_of_permits = length(Category))

ggplot(counts, aes(x = Category, y = count_of_permits )) +
  geom_bar(stat="identity")+ 
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

```

### ggplot2 basics

Before we go further, I want to point out some basic properties of ggplot2, just to give you a sense of how it is working. This will make more sense in a few weeks, so come back here to remind yourself. We'll do just a bit a basics, and then move on to making more graphs, by copying and pasting.

The ggplot function makes use of so-called layers. Layers you say? What are these layers? Well, it draws things from the bottom up. It lays down one layer of graphics, then you can keep adding on top, drawing more things. So the idea is something like: Layer 1 + Layer 2 + Layer 3, and so on. If you want Layer 3 to be Layer 2, then you just switch them in the code.

Here is a way of thinking about ggplot code

```{r, echo=TRUE,eval=FALSE}
ggplot(name_of_data, aes(x = name_of_x_variable, y = name_of_y_variable)) +
    geom_layer()+
    geom_layer()+
    geom_layer()

```

What I want you to focus on in the above description is the $+$ signs. What we are doing with the plus signs is adding layers to plot. The layers get added in the order that they are written. If you look back to our previous code, you will see we add a `geom_bar` layer, then we added another layer to change the rotation of the words on the x-axis. This is how it works.

BUT WAIT? How am I supposed to know and remember what to add? This is nuts! We know. You're not supposed to know just yet, how could you? We'll give you lots of examples where you can copy and paste, and they will work. That's how you'll learn for now. If you really want to read the [help manual](https://ggplot2.tidyverse.org/reference/index.html) you can do that too. It's on the ggplot2 website. This will become useful after you already know what you are doing, but before that, it will probably just seem very confusing. However, it is pretty neat to look and [see all of the different things you can do](https://exts.ggplot2.tidyverse.org/gallery/), it's very powerful.

For now, let's the get the hang of adding things to the graph that let us change some stuff we might want to change. For example, how do you add a title? Or change the labels on the axes? Or add different colors, or change the font-size, or change the background? You can change all of these things by adding different lines to the existing code.

#### ylab() changes y label

The last graph had `count_of_permits` as the label on the y-axis. That doesn't look right. ggplot2 automatically took the label from the column, and made it be the name on the y-axis. We can change that by adding `ylab("what we want")`. We do this by adding a $+$ to the last line, then adding `ylab()`

```{r 1categoryC,echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits")
```

#### xlab() changes x label

Let's slightly modify the x label too:

```{r 1categoryD, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film")
```


#### ggtitle() adds title

Let's give our graph a title

```{r 1categoryE, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category")
```

#### color adds color

Let's make the bars different colors. To do this, we add new code to the inside of the `aes()` part:

```{r 1categoryF, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, color=Category )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category")
```

#### fill fills in color

Let's make the bars different colors. To do this, we add new code to the inside of the `aes()` part...Notice I've started using new lines to make the code more readable. 

```{r 1categoryG, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category")
```

#### get rid of the legend

Sometimes you just don't want the legend on the side, to remove it add 

`theme(legend.position="none")`

```{r 1categoryH, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category") +
  theme(legend.position="none")
```

#### theme_classic() makes white background

The rest is often just visual preference. For example, the graph above has this grey grid behind the bars. For a clean classic no nonsense look, use `theme_classic()` to take away the grid.

```{r 1categoryI, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category") +
  theme(legend.position="none") +
  theme_classic()
```

#### Sometimes layer order matters

Interesting, `theme_classic()` is misbehaving a little bit and incorrectly renders the axis labels and reintroduces the legend. It looks like we have some of our layers out of order, let's re-order. I just moved `theme_classic()` to just underneath the `geom_bar()` line. Now everything get's drawn properly.

```{r 1categoryJ, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category") +
  theme(legend.position="none") 
  

```

#### Font-size

Changing font-size is often something you want to do. ggplot2 can do this in different ways. I suggest using the `base_size` option inside `theme_classic()`. You set one number for the largest font size in the graph, and everything else gets scaled to fit with that that first number. It's really convenient. Look for the inside of `theme_classic()`

```{r 1categoryK, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme_classic(base_size = 15) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category") +
  theme(legend.position="none") 
  

```
or make things small... just to see what happens

```{r 1categoryL, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category") +
  theme(legend.position="none") 
  

```

#### ggplot2 summary

That's enough of the ggplot2 basics for now. You will discover that many things are possible with ggplot2. It is amazing. We are going to get back to answering some questions about the data with graphs. But, now that we have built the code to make the graphs, all we need to do is copy-paste, and make a few small changes, and boom, we have our graph.

### More questions about NYC films

#### What are the sub-categories of films?
 
Notice the `nyc_films` data frame also has a column for `SubCategoryName`. Let's see what's going on there with a quick plot.

```{r 1subcategory, echo=TRUE,eval=TRUE}

# get the counts (this is a comment)
# comments are really useful: R will ignore them, but they can explain to a reader 
# what is going on in the code

counts <- nyc_films %>%
          group_by(SubCategoryName) %>%
          summarize(count_of_permits = length(SubCategoryName))

# make the plot

ggplot(counts, aes(x = SubCategoryName, y = count_of_permits, 
                   color=SubCategoryName, 
                   fill= SubCategoryName )) +
  geom_bar(stat="identity") + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Sub-category of film") +
  ggtitle("Number of Film permits in NYC by Sub-category") +
  theme(legend.position="none") 

```

I guess "episodic series" are the most common. Using a graph like this gave us our answer super fast.

#### Categories by different Boroughs

Let's see one more really useful thing about ggplot2. It's called `facet_wrap()`. It's an ugly word, but you will see that it is very cool, and you can do next-level-super-hero graph styles with `facet_wrap` that other people can't do very easily. 

Here's our question. We know that some films are made in different Boroughs, and that same films are made in different categories, but do different Boroughs have different patterns for the kinds of categories of films they request permits for? Are their more TV shows in Brooklyn? How do we find out? Watch, just like this:

```{r 1facetwrap, echo=TRUE,eval=TRUE}

# get the counts

counts <- nyc_films %>%
          group_by(Borough,Category) %>%
          summarize(count_of_permits = length(Category))

# make the plot

ggplot(counts, aes(x = Category, y = count_of_permits, 
                   color=Category, 
                   fill= Category )) +
  geom_bar(stat="identity") + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Category of film") +
  ggtitle("Number of Film permits in NYC by Category and Borough") +
  theme(legend.position="none") +
  facet_wrap(~Borough, ncol=3)

```

We did two important things. First we added `Borough` and `Category` into the `group_by()` function. This automatically gives separate counts for each category of film, for each Borough. Then we added `facet_wrap(~Borough, ncol=3)` to the end of the plot, and it automatically drew us 5 different bar graphs, one for each Borough! That was fast. Imagine doing that by hand.

The nice thing about this is we can switch things around if we want. For example, we could do it this way by switching the `Category` with `Borough`, and facet-wrapping by Category instead of Borough like we did above. Do what works for you.


```{r 1facetwrap2, echo=TRUE,eval=TRUE}
ggplot(counts, aes(x = Borough, y = count_of_permits, 
                   color=Borough, 
                   fill= Borough )) +
  geom_bar(stat="identity") + 
  theme_classic(base_size = 10) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of Film Permits") + 
  xlab("Borough") +
  ggtitle("Number of Film permits in NYC by Category and Borough") +
  theme(legend.position="none") +
  facet_wrap(~Category, ncol=5)

```

### Gapminder Data

[Gapminder](https://www.gapminder.org) is an organization that collects some really interesting worldwide data. They also make cool visualization tools for looking at the data. There are many neat examples, and they have visualization tools built right into their website that you can play around with [https://www.gapminder.org/tools/](https://www.gapminder.org/tools/). That's fun to check out.

There is also an R package called `gapminder`. When you install this package, it loads in some of the data from gapminder, so we can play with it in R.

If you don't have the gapminder package installed, you can install it by running this code

```{r, eval=FALSE,echo=TRUE}
install.packages("gapminder")
```

Once the package is installed, you need to load the new library, like this. Then, you can put the `gapminder` data into a data frame, like we do here: `gapminder_df`.

```{r, echo=TRUE,eval=TRUE}
library(gapminder)
gapminder_df<-gapminder
```

#### Look at the data frame

You can look at the data frame to see what is in it, and you can use `summarytools` again to view a summary of the data.

```{r, echo=TRUE,eval=FALSE}
view(dfSummary(gapminder_df))
```

There are 1704 rows of data, and we see some columns for country, continent, year, life expectancy, population, and GDP per capita.

### Asking questions about the gapminder data

We will show you how to graph some the data to answer a few different kinds of questions. Then you will form your own questions, and see if you can answer them with ggplot2 yourself. All you will need to do is copy and paste the following examples, and change them up a little bit

#### Life expectancy histogram

How long are people living all around the world according to this data set? There are many ways we could plot the data to find out. The first way is a histogram. We have many numbers for life expectancy in the column `lifeExp`. This is a big sample, full of numbers for 142 countries across many years. It's easy to make a histogram in ggplot to view the distribution:

```{r 1gapminder, echo=TRUE,eval=TRUE}
ggplot(gapminder_df, aes(x=lifeExp))+
  geom_histogram(color="white")
```

See, that was easy. Next, is a code block that adds more layers and settings if you wanted to modify parts of the graph:

```{r 1gapminderB, echo=TRUE,eval=TRUE}

ggplot(gapminder_df, aes(x = lifeExp)) +
  geom_histogram(color="white")+ 
  theme_classic(base_size = 15) +
  ylab("Frequency count") + 
  xlab("Life Expectancy") +
  ggtitle("Histogram of Life Expectancy from Gapminder")

```

The histogram shows a wide range of life expectancies, from below 40 to just over 80. Histograms are useful, they can show you what kinds of values happen more often than others.

One final thing about histograms in ggplot. You may want to change the bin size. That controls how wide or narrow, or the number of bars (how they split across the range), in the histogram. You need to set the `bins=` option in `geom_histogram()`.

```{r 1gapminderC, echo=TRUE,eval=TRUE}

ggplot(gapminder_df, aes(x = lifeExp)) +
  geom_histogram(color="white", bins=50)+ 
  theme_classic(base_size = 15) +
  ylab("Frequency count") + 
  xlab("Life Expectancy") +
  ggtitle("Histogram of Life Expectancy from Gapminder")

```

See, same basic patter, but now breaking up the range into 50 little equal sized bins, rather than 30, which is the default. You get to choose what you want to do.

#### Life expectancy scatterplot

We can see we have data for life expectancy and different years. So, does worldwide life expectancy change across the years in the data set? As we go into the future, are people living longer?

Let's look at this using a scatter plot. We can set the x-axis to be year, and the y-axis to be life expectancy. Then we can use `geom_point()` to display a whole bunch of dots, and then look at them. Here's the simple code:

```{r 1scatterplot, echo=TRUE,eval=TRUE}
ggplot(gapminder_df, aes(y= lifeExp, x= year))+
  geom_point()
```

Whoa, that's a lot of dots! Remember that each country is measured each year. So, the bands of dots you see, show the life expectancies for the whole range of countries within each year of the database. There is a big spread inside each year. But, on the whole it looks like groups of dots slowly go up over years.

#### One country, life expectancy by year

I'm (Thomas) from The Netherlands, so maybe I want to know if life expectancy for Dutch people is going up over the years. To find out the answer for one country, we first need to split the full data set, into another smaller data set that only contains data for The Netherlands. In other words, we want only the rows where the word "Netherlands" is found in the `country` column. We will use the `filter` function from `dplyr` for this:

```{r 1scatterB, echo=TRUE,eval=TRUE}
# filter rows to contain Canada

smaller_df <- gapminder_df %>% 
                 filter(country == "Netherlands")

# plot the new data contained in smaller_df

ggplot(smaller_df, aes(y= lifeExp, x= year))+
  geom_point()

```

I would say things are looking good for Dutch people, their life expectancy is going up over the years!

#### Multiple countries scatterplot

What if we want to look at a few countries altogether. We can do this too. We just change how we filter the data so more than one country is allowed, then we plot the data. We will also add some nicer color options and make the plot look pretty. First, the simple code:

```{r 1scatterC, echo=TRUE,eval=TRUE}

# filter rows to contain countries of choice

smaller_df <- gapminder_df %>% 
                 filter(country %in% c("Netherlands","France","Brazil") == TRUE)

# plot the new data contained in smaller_df

ggplot(smaller_df, aes(y= lifeExp, x= year, group= country))+
  geom_point()

```

Nice, we can now see three sets of dots, but which are countries do they represent? Let's add a legend, and make the graph better looking.

```{r 1scatterD, echo=TRUE,eval=TRUE}
ggplot(smaller_df,aes(y= lifeExp, x= year, 
                      group= country, color = country)) +
  geom_point()+ 
  theme_classic(base_size = 15) +
  ylab("Life Expectancy") + 
  xlab("Year") +
  ggtitle("Life expectancy by year for three countries")

```

#### geom_line() connecting the dots

We might also want to connect the dots with a line, to make it easier to see the connection! Remember, ggplot2 draws layers on top of layers. So, we add in a new `geom_line()` layer.

```{r 1scatline, echo=TRUE,eval=TRUE}
ggplot(smaller_df,aes(y= lifeExp, x= year, 
                      group= country, color = country)) +
  geom_point()+ 
  geom_line()+
  theme_classic(base_size = 15) +
  ylab("Life Expectancy") + 
  xlab("Year") +
  ggtitle("Life expectancy by year for three countries")

```

### Generalization Exercise

The following generalization exercises and writing assignment can be answered in the R Markdown document for this lab. Complete your work in that document and "knit" the file. Hand in your final product (knitting should generate an .html file) to Canvas. If you are unable to knit, submit the .Rmd file instead. 

Use the code from above to attempt to solve the extra things we ask you do for this assignment. Your generalization exercises are as follows:

1. Make a graph plotting Life Expectancy by year for the five continents, using the `continent` factor. Make sure you change the title as well.

2. Make a graph plotting GDP per capita by year for the USA, Canada, and Mexico. Use the `gdpPercap` column for the GDP per capita data.

3. Make a new graph plotting anything you are interested in using the gapminder dataset. It just needs to be a plot that we have not given an example for.

### Writing assignment

Describe what histograms are, how to interpret them, and what they are useful for. You should answer each of these questions:

a. What do the bars on a histogram represent?
b. How many bars can a histogram have?
c. What do the heights of the bars tell you?
d. What is on the x-axis and y-axis of a histogram?
e. What does the tallest bar on a histogram tell you?
f. What does the shortest bar on a histogram tell you?
g. What are some uses for histograms, why would you want to look at a histogram of some numbers that you collected?
h. Imagine you had two histograms, one was very wide and spread out, the other was very narrow with a very tall peak. Which histogram would you expect to contain more consistent numbers (numbers that are close to each other) and explain why.

