---
title: "Lab 4 - Chance and Probability Theory"
author: "Student Name"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = FALSE, error = TRUE, message = FALSE, warning = TRUE, tidy = FALSE)
```

Welcome to the template file of the third lab! Today is all about chance and probability theory. Start this lab by reading Section 4.2 of the lab manual. Make sure you follow along with the code examples given in the lab manual. 

```{r}
# Run this code chunk to make sure the tidyverse is loaded in your R session. 
library(tidyverse) 
```

# Correlation and random chance
After reading Section 4.2, complete the exercises below. 

## Chance exercises
1. Estimate the range (minimum and maximum) of correlations that could occur by chance between two normally distributed random variables with n = 10.
2. Estimate the range (minimum and maximum) of correlations that could occur by chance between two normally distributed random variables with n = 25.

Modify the following code to answer question 1 and 2.  

```{r}
saved_value <- c() # create empty variable
for (i in 1:5) { # loop over code between {} 5 times
  x <- rnorm(n=10, mean=0, sd=5) # generate normally distributed random variables
  y <- rnorm(n=10, mean=0, sd=5)
  saved_value[i] <- sum(x*y) # save the sum of the product of x*y in saved_value
}

min_value <- min(saved_value)
max_value <- max(saved_value)

print(min_value)
print(max_value)

```

3. What proportion of simulated correlations (from question 1 and 2) is smaller than -0.3 and larger than +0.3 for n = 10 and n = 25? To calculate the proportion use the smaller than (`<`) and greater than (`>`) operators on your vector of correlations like so: `mean(saved_value < 0.3)`

```{r}

```

4. The histogram in the lab manual shows the correlations produced by random chance for n = 10. The red lines are set at -0.3 and +0.3. The text indicates the proportion of correlations outside the red lines. Try to recreate the histogram from the lab manual as closely as possible for n = 25.

```{r}

```

Remember, programming is an iterative process, you almost never get to the solution in the first go. Make liberal use of the help function in R, Google, the tidyverse website and/or the cheatsheets and ask help from your classmates / the tutor when stuck. There are multiple ways to get to the plot below. 

After completing these exercises, continue the lab by reading Section 4.3 of the lab manual.

# Generating data in R

## `sample` and `binom()` exercises
1. Why are you unable to run the following bit of code? What could you do to fix this? Does this change the nature of your sampling procedure?
```{r,eval=F,echo=T}
sample(c(1:10), 20)
```

2. Simulate the amount of sixes you expect to roll when rolling 1 dice 10,000 times. Also simulate the amount of fives **and** sixes you expect to roll when rolling 10 dice 10,000 times? (hint: use `rbinom()` to simulate 10,000 dice rolls and `sum()` to calculate the amount of successes)

```{r}

```

3. What is the probability of obtaining exactly 4 heads when flipping 10 fair coins? And what is the probability to obtain **at most** 4 heads (so 0, 1, 2, 3, or 4 heads)? Cf. [the textbook](https://thomashulst.github.io/quantrma/chance.html#working-with-the-binomial-distribution-in-r) when unsure how to use the `binom()` functions.

```{r}

```

4. What is the probability to obtain **at least** 4 heads (so 4, 5, 6, 7, 8, 9, or 10 heads) when flipping 10 fair coins? You can calculate this probability in two ways, either by using `pbinom()` with an additional argument, or by using your answers from question 3. 

```{r}

```

After completing these exercises, continue the lab by reading Section 4.4 of the lab manual.

## `norm()` exercises
Complete the following exercises after reading Section 4.4 of the lab manual.

Run the following bit of code, which samples 20 random numbers from a normal distribution, a couple of times and look at the results.

```{r}
set.seed(123)
some_numbers <- rnorm(20,50,25) # 20 numbers, mean = 50, s.d. = 25
print(some_numbers)
```

1. What do you think the function `set.seed()` does? When do you think this could be useful?

2. Suppose the mean of a normal distribution is 25 ($\mu = 25$) and the standard deviation is 3 ($\sigma=3$). Calculate the probability of obtaining a value between 22 and 28 using R. (hint: `pnorm()`)

```{r}

```

3. Based on what you know about the standard normal distribution, could you answer question 2 without calculations?

4. Use R to calculate the probability of obtaining a value **higher than 29.5** for the normal distribution with $\mu = 25$ and $\sigma=3$. (hint: `pnorm()`)

```{r}

```

Continue the lab by reading Section 4.5 of the lab manual.

## z-score exercises
Complete the exercises below after reading Section 4.5 of the lab manual.

1. Create a table containing the mean and standard deviation for this sample of scores. Also produce a histogram of the grades. How does the distribution of grades look with a binwidth of 5?
```{r}

```

2. Transform each student’s score into a Z-score. Now, plot the frequency histogram  of this Z-score distribution with a binwidth of 0.5. Compare it to the raw score distribution. How are they the same? How are they different?

```{r}

```

3. Imagine you are a student in this class who received a 90 on this exam. However, the Professor has decided to grade on a curve, such that only the top 10% of the class receives an A. Calculate the z-score that corresponds to a raw score of 90 on this exam. Will you get an A with this grade? Why or why not?
```{r}

```


When you have completed all exercises and are happy with your progress today, please knit your document and submit it to Canvas. If you finish before the time is up, start with the required readings of Week 5, work on your assignment, or help out your fellow students.
