---
title: "Lab 2 - Describing Data"
author: "Student Name"
date: "`r Sys.Date()`"
output:
  word_document:
    toc: yes
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = FALSE, error = TRUE, message = FALSE, warning = TRUE, tidy = TRUE)
```

Welcome to the template file of the second lab! Today is all about describing data.

We are going to introduce a lot of new programming concepts this week very quickly and you might feel overwhelmed or unsure you fully grasp the new concepts. This is completely normal when learning a new programming language. We start with simply copy/pasting and dissecting R commands as a way to familiarize you with the language. You will see that by doing this you can very quickly learn to answer interesting questions about a data set, just from changing a couple of minor things in the code examples. In our experience, this works better than expecting you to code something from scratch at the start of learning a new language. As always, there are several ways to get help when coding in R:

1. Use the ?help or ??help functions
2. Google your question
3. Ask a fellow student or your tutor for additional explanations
4. Have a look at the R cheatsheets on Canvas or [tidyverse website](https://www.tidyverse.org)

Start this lab by reading the learning goals and Section 2.2 of the manual. Make sure you really follow along with the code examples given, either in the console, or (even better) by including the code examples in a code chunk in this R Markdown document and running it from there. Like so:

```{r}
# This is the first code example from the lab manual and will load the Film_Permits.csv file into R
library(tidyverse)
nyc_films <-read_csv("data/Film_Permits.csv")
```

# Loading data 

## Data exercises
Complete the following exercises after reading Section 2.2 of the lab manual. 

1. What are the last 6 entries of the `Category` column in `nyc_films`? 
```{r}

```

2. What are the last 10 entries of the `EventType` column in `nyc_films`? (hint: use `tail()` with the `n=` argument)
```{r}

```

3. You can also load data directly from the internet using `read_csv()`. Use `read_csv()` to load the data from https://github.com/thomashulst/quantrma_lab/raw/master/data/Film_Permits.csv and assign it to the variable `nyc_films2`.
```{r}

```

Continue this lab by reading Section 2.3 of the lab manual. You don't have to read Section 2.3 in its entirety, just Section 2.3.1, but what you should read is clearly indicated in the manual. Make sure to follow along with the code examples given. 

# Making graphs

## Plot exercises
After reading section 2.3 of the lab manual, complete to following exercises:

1. Explain in your own words what is happening in the code block below. What does the `%>%` operator indicate?
```{r}
counts <- nyc_films %>% 
  group_by(Category) %>% 
  summarize(count_of_permits = length(Category))
```
2. The first argument of `ggplot()` should always be a data argument. In what type of form should the data be provided? (hint: `?ggplot`)
3. What function can we use to set the title of a plot? Use the ggplot [reference manual](https://ggplot2.tidyverse.org/reference/index.html), the additional materials, or Google to answer this question.

Continue this lab by reading Section 2.4 of the lab manual and follow along with the code examples given.

# Gapminder dataset

## Gapminder exercises
After reading section 2.4 of the lab manual, complete the exercises below. All code exercises can be solved by copy/pasting code from prior examples and changing just 1 or 2 things.  

1. Make a histogram of GDP per capita in the Gapminder dataset. (hint: see the histogram of life expectancy)
```{r}


```

2. Make a time series plot of GDP per capita by year for the Brazil, Canada, and Mexico using the code above. Make sure to update the title and labels as well.
```{r}


```

3. Make a time series plot of Life Expectancy by year for the continents (hint: use the `group_by()` and `summarise()` functions from grouping and summarizing the NYC film permits).
```{r}


```

Continue this lab by reading Section 2.5 of the lab manual and follow along with the code examples given.

# Using numbers to describe data

## Descriptive exercises
After reading section 2.5 of the lab manual, complete the exercises below. All code exercises can be solved by copy/pasting code from prior examples and changing just 1 or 2 things.  

1. What is the mean, standard deviation, minimum and maximum GDP per capita for all the gapminder data (across all the years and countries)? Print the values in a table with your own caption.
```{r}

```

2. What is the mean, median, variance, and length of the life expectancy variable for each continent in 2007? Print the values in a table with your own caption.
```{r}

```

3. Answer the following questions in your own words:

a. Define the mode.
Answer here.

b. Explain what would need to happen in order for a set of numbers to have two modes
Answer here.
  
c. Define the median 
Answer here.
  
d. Define the mean 
Answer here.
  
e. Define the range
Answer here.
  
f. When calculating the variance, explain what the difference scores represent
Answer here.
  
g. Explain why the difference scores are squared when calculating the variance
Answer here.
  
h. If one set of numbers had a standard deviation of 5, and another had a standard deviation of 10, which set of numbers would have greater variance? Explain why.
Answer here.

When you have completed all exercises and are happy with your progress today, please knit your document and submit it to Canvas. If you finish before the time is up, you can help out your fellow students, start with the required readings of Week 3, or work on your assignment. If you haven't done so already, make sure you've [signed up for an assignment group](https://canvas.eur.nl/courses/31825/assignments/118509).