---
title: "Lab 4 - Chance and Probability Theory"
author: "Student Name"
date: "xx/xx/xx"
output:
  word_document:
    toc: yes
  html_document:
    toc: yes
    toc_float: yes
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = FALSE, error = TRUE, message = FALSE, warning = TRUE, tidy = TRUE)
```

You can keep notes about the lab here and execute code like in the code block below:

```{r}
# Type or copy/paste R code in an R code block to have the code executed and the output printed in your final (knitted) document.
10+20
```

# Part one: correlation and random chance

## Chance exercises

#### 1. Estimate the range (minimum and maximum) of correlations that could occur by chance between two variables with n = 10.

```{r}

saved_value <- c() #make an empty variable
for (i in 1:5) {
  x <- runif(n=10, min=1, max=10)
  y <- runif(n=10, min=1, max=10)
  saved_value[i] <- sum(x*y)
}

min(saved_value)
max(saved_value)

```


#### 2. Estimate the range (minimum and maximum) of correlations that could occur by chance between two variables with n = 25.

```{r}

saved_value <- c() #make an empty variable
for (i in 1:5) {
  x <- runif(n=10, min=1, max=10)
  y <- runif(n=10, min=1, max=10)
  saved_value[i] <- sum(x*y)
}

min(saved_value)
max(saved_value)

```

#### 3. What proportion of correlations (from question 1 and 2) is smaller than -0.3 and larger than +0.3 for n = 10 and n = 25? Create a  histogram of the correlations for n = 10 and n = 25 and draw a vertical line at -0.3 and +0.3. Print the found proportions on the plots. Use `ggarrange()` from the `ggpubr` library to plot the histograms alongside each other in one figure. 

```{r,tidy=F}
library(ggpubr)
# see ?ggarrange() 
# in short, you need to assign your ggplots to a variable like so:
# plot1 <- ggplot(...)
# plot2 <- ggplot(...)
# ggarrange(plot1, plot2)

```

# Part two: generating data in R

## `sample` and `binom()` exercises

#### 1. Why are you unable to run the following bit of code? What could you do to fix this? Does this change the nature of your sampling procedure?
```{r}
sample(c(1:10), 20)
```
#### 2. How many sixes do you expect to roll when rolling 1 dice 10,000 times? And how many fives **and** sixes do you expect to roll when rolling 10 dice 10,000 times?
```{r}

```

#### 3. What is the probability of obtaining exactly 4 heads when flipping 10 fair coins? And what is the probability to obtain **at least** 4 heads (so 4, 5, 6, 7, 8, 9 or 10 heads)? Cf. [the textbook](https://thomashulst.github.io/quantrma/chance.html#working-with-the-binomial-distribution-in-r) when unsure how to use the `binom()` functions.
```{r}

```

## `norm()` exercises
Run the following bit of code, which samples 20 random numbers from a normal distribution, a couple of times and look at the results. 

#### 1. What do you think the function `set.seed()` does? When do you think this could be useful?
```{r}
set.seed(123)
some_numbers <- rnorm(20,50,25) # 20 numbers, mean = 50, s.d. = 25
print(some_numbers)
```

#### 2. Suppose the mean of a normal distribution is 25 ($\mu = 25$) and the standard deviation is 3 ($\sigma=3$). Calculate the probability of obtaining a value between 22 and 28 using R.

```{r}

```

#### 3. Based on what you know about the standard normal distribution, could you have figured this out without calculations?

#### 4. Use R to calculate the probability of obtaining a value higher than 29.5 for this normal distribution.

```{r}

```

## z-score exercises
[Right click and download this](https://github.com/thomashulst/quantrma_lab/blob/master/data/spssdata/StatsGrades.sav?raw=true) SPSS file containing 49 students’ exam grades (let’s say it’s the final exam for a statistics class). You have likely never worked with an SPSS (.sav) file before, but still, with a quick Google search, you should be able to find a library and function to load the data into R. Remember, you can install packages using `install.packages()` and load a library with `library()`. Make sure the .sav is in your working directory so you can load it into R. 

Once you have successfully loaded the data into R:

#### 1. Create a table containing the mean and standard deviation for this sample of scores. Also produce a frequency histogram of the grades. How does the distribution of grades look?
```{r}

```

#### 2. Transform each student’s score into a Z-score. Now, plot the frequency histogram of this Z-score distribution. Compare it to the raw score distribution. How are they the same? How are they different?

```{r}

```

#### 3. Imagine you are a student in this class who received a 90 on this exam. However, the Professor has decided to grade on a curve, such that only the top 10% of the class receives an A (this professor only gives whole grades, no minuses or pluses). Calculate the z-score that corresponds to a raw score of 90 on this exam. Will you get an A with this grade? Why or why not?
```{r}

```
