---
title: "Lab 2 - Describing Data"
author: "Student Name"
date: "xx/xx/xx"
output:
  html_document:
    toc: yes
    toc_float: yes
  word_document:
    toc: yes
---

```{r setup,include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, cache = FALSE, error = TRUE, message = TRUE, warning = TRUE, tidy = TRUE)
```

You can keep notes about the lab here and execute code like in the code block below:

```{r}
# Type or copy/paste R code in an R code block to have the code executed and the output printed in your final (knitted) document.
10+20
```

# Part one: describing data using graphs 

## Data exercises

#### 1. What are the last 6 entries of the `Category` column in `nyc_films`? 
```{r}

```

#### 2. What are the last 10 entries of the `EventType` column in `nyc_films`? 
```{r}

```

#### 3. Use `fread()` to only load the `EventID` column and assign it to a variable named `nyc_films_eventid` (hint: ?fread()).
```{r}

```

## Graphing exercises

#### 1. Make a graph plotting Life Expectancy by year for the five continents, using the `continent` factor. Make sure you change the title as well.
```{r}

```

#### 2. Make a graph plotting GDP per capita by year for the USA, Canada, and Mexico. Use the `gdpPercap` column for the GDP per capita data.
```{r}

```

#### 3. Make a new graph plotting anything you are interested in using the gapminder dataset. It just needs to be a plot that we have not given an example for.
```{r}

```

# Part two: describing data using numbers

## Descriptive exercises

#### 1. What is the mean, standard deviation, minimum and maximum life expectancy for all the gapminder data (across all the years and countries) (hint: do not use `group_by`)
```{r}

```

#### 2. What is the mean, standard deviation, minimum and maximum life expectancy for all of the continents in 2007, the most recent year in the dataset (hint: add another pipe using `filter(year==2007) %>%`).
```{r}

```

#### 3. Answer the following questions in full sentences using simple plain language:

##### a. Define the mode.
Answer here.

##### b. Explain what would need to happen in order for a set of numbers to have two modes
Answer here.
  
##### c. Define the median 
Answer here.
  
##### d. Define the mean 
Answer here.
  
##### e. Define the range
Answer here.
  
##### f. When calculating the standard deviation, explain what the difference scores represent
Answer here.
  
##### g. Explain why the difference scores are squared when calculating the standard deviation
Answer here.
  
##### h. If one set of numbers had a standard deviation of 5, and another had a standard deviation of 10, which set of numbers would have greater variance? Explain why.
Answer here.

When you have completed all exercises and are happy with your progress today, please knit your document and submit it to Canvas. If you finish before the time is up, start with the required readings of Week 3 or help out your fellow students. 